---
title: "R + Greenplum + Excel Tutorial"
author: "Joshua Ruf"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

```{r setup}
library(fidelis) # also attaches data.table
library(openxlsx) # R and Excel integration
library(tidyverse)
library(formattable) # for coloring data.table ouputs
```

This vignette outlines how we can use R to connect to the PostgreSQL Greenplum database, integrate R with our Excel workfow, and create pretty html based reports. As an example, we will be creating a report that compares Instututional claims with Professional claims across a number of LOB, by looking at the paid dates.<br>

Before we begin, set the pagewidth of the html output with this function below, be sure to set `results='asis'` in the Rmarkdown chunk.

```{r, results='asis'}
fidelis::setup_pagewidth(1200)
```


To Begin we need a few things:<br>
1. An excel template file with 6 sheets:<br>
    a) INSTITUTIONAL, a formula driven summary tab<br>
    b) INSTITUTIONAL_RAW, an invisible sheet to dump raw data<br>
    c) PROFESSIONAL, a formula driven summary tab<br>
    d) PROFESSIONAL_RAW, an invisible sheet to dump raw data<br>
    e) LOB, an invisible sheet to dump product names to populate dropdown list. Note: cannot be empty initially as the dropdown formula won't work<br>
    f) WEEKS_RAW, an invisible sheet to dump how many fridays each month has<br>
2. A secret password file created with `fidelis::create_secret()`, allowing us to use the ask=F option in `fidelis::greenplum_connect()`<br>
3. Make sure the template file is not open while you run this script!

### Formula-based Excel Document

```{r}
fidelis::greenplum_connect(ask=F)
# create greenplum connection object, see ?fidelis::greenplum_connect() for more information
# once we've created a connection object with this function, query() automatically uses that connection to query

wb <- openxlsx::loadWorkbook('Weekly_Cost_Utilization_by_Paid_Date_TEMPLATE.xlsx')
# create a wb workbook binding object for the template file
```

First, we'll add the WEEKS_RAW tab.<br>

The `query()` function in the fidelis package is dynamic, meaning that you can pass r objects as parameters into the query. To make this explicit, I will create an R object that is the minimum effective period date to limit the queries.


```{r}
min_effper <- '1/1/2016'

fridays <- fidelis::query(

  "
  SELECT date_trunc('month', clcl_paid_dt) AS actper
    ,count(*) AS fridays
  FROM (
  	SELECT DISTINCT clcl_paid_dt
  	FROM ods_facets.cmc_clcl_claim
  	WHERE clcl_paid_dt >= %min_effper%) AS a
  WHERE to_char(clcl_paid_dt, 'day') LIKE 'friday%'
  GROUP BY 1
  ;
  ",
  min_effper = min_effper) %>%
  dplyr::mutate(actper = as.Date(actper, format = '%m/%d/%Y'))
# return the results of the above query and convert dates to normal date format

openxlsx::removeWorksheet(wb, 'WEEKS_RAW') # drop old WEEKS_RAW tab
openxlsx::addWorksheet(wb, 'WEEKS_RAW', visible = F) # create new hidden WEEKS_RAW tab
openxlsx::writeData(wb, 'WEEKS_RAW', fridays) # dump the data.frame fridays in the WEEKS_RAW tab of wb
```

Next, we'll add the INSTITUTIONAL_RAW and PROFESSIONAL_RAW tabs.

```{r}
fidelis::query("

  DROP TABLE IF EXISTS final;
  CREATE temp TABLE final AS
  SELECT date_trunc('month', clcl_paid_dt) AS actper,
    clcl_cl_sub_type,
    product,
    mmcor_desc,
    sum(cdml_paid_amt) AS paid,
    count(DISTINCT clcl_id) AS util
  FROM sandbox.mmcor_final_medecon_gl as a
  where clcl_paid_dt >= %min_effper%
    and product not in ('FIDA', 'FHP')
    and product is not NULL
  GROUP BY 1,2,3,4
  ;
  ",
  min_effper = min_effper
)
# create a temp table, final, to pull from in the next query

institutional <- fidelis::query(

  "
  select actper
    ,null as placeholder1
    ,null as placeholder2
    ,mmcor_desc
    ,'All' as placeholder3
    ,paid
    ,util
    ,product
  from final
  where clcl_cl_sub_type = 'H'
  ;
  "
  ) %>%
  dplyr::mutate(actper = as.Date(actper, format = '%m/%d/%Y'))

openxlsx::removeWorksheet(wb, 'INSTITUTIONAL_RAW')
openxlsx::addWorksheet(wb, 'INSTITUTIONAL_RAW', visible = F)
openxlsx::writeData(wb, 'INSTITUTIONAL_RAW', institutional)

professional <- fidelis::query(

  "
  select actper
    ,null as placeholder1
    ,null as placeholder2
    ,mmcor_desc
    ,'ALL' as placeholder3
    ,paid
    ,util
    ,product
  from final
  where clcl_cl_sub_type = 'M'
  ;
  "
  ) %>%
  dplyr::mutate(actper = as.Date(actper, format = '%m/%d/%Y'))

openxlsx::removeWorksheet(wb, 'PROFESSIONAL_RAW')
openxlsx::addWorksheet(wb, 'PROFESSIONAL_RAW', visible = F)
openxlsx::writeData(wb, 'PROFESSIONAL_RAW', professional)
```

```{r}
lob <- unique(professional$product)

openxlsx::removeWorksheet(wb, 'LOB')
openxlsx::addWorksheet(wb, 'LOB', visible = F)
openxlsx::writeData(wb, 'LOB', lob, colNames = F) # do not add the column name
```

Save the updated excel workbook to a new file, named based on the current date

```{r}
openxlsx::saveWorkbook(wb, overwrite = TRUE,
             file = sprintf("Weekly_Cost_Utilization_by_Paid_Date_%s.xlsx", gsub('-', '', Sys.Date())))
```

Take a look at the file: `r sprintf("Weekly_Cost_Utilization_by_Paid_Date_%s.xlsx", gsub('-', '', Sys.Date()))`, remembering that there are hidden worksheets.

### Excel Data Dump

Now suppose that we are in need of the raw data for the professional claims, with each LOB in a different tab. This can also be accomplished with the openxlsx package.

```{r}
professional_list <- split(professional,
                           f = professional$product,
                           drop = T)
# create a list of data.frames split by product; drop=T means we'll drop the provider variable in each data.frame in the list

openxlsx::write.xlsx(professional_list,
           file = sprintf("professional_raw_%s.xlsx", gsub('-', '', Sys.Date())))
```

Take a look at the workbook: `r sprintf("professional_raw_%s.xlsx", gsub('-', '', Sys.Date()))`.


### HTML Output

Finally, let's say we want to bypass Excel altogether and instead create an html-based table, perhaps even color-coded. As an example we'll create a series of tables for each LOB for only professional claims. Part of the usefulness of this approach is the ability to loop through each LOB.

```{r}
final_effper <- lubridate::add_with_rollback(e1 = lubridate::round_date(Sys.Date(), 'months'), e2 = months(-4))
# lets look back such that the final effper still gives three complete months between now and the final day in that month.

months <- 3 # use a three month rolling window

dates_df <- final_effper %>%
  fidelis::set_dates() %>%
  .[, c(1, 1 + months/3), with = F]
# assigns each month in a 24 month period to either prior, current, or NULL

current <- dates_df[col == 'current', effper]
# current period

prior <- dates_df[col == 'prior', effper]
# prior period

cos_names <- data.table::fread('cos_names.csv')
# data.table with sandbox mmcor_desc names as well as prettied ones used in final reports

fridays <- fridays %>%
  data.table::as.data.table() %>%
  .[actper %in% c(current, prior)] %>%
  .[, col := ifelse(actper %in% current, 'current', 'prior')] %>%
  .[, .(fridays = sum(fridays)), by = col]
# aggregate fridays for the current and prior period


df <- professional %>%
  merge(cos_names[subtotal==1, .(cat, category)],
        by.x = 'mmcor_desc',
        by.y = 'cat') %>% # clean the mmcor_desc names
  .[, -grep('placeholder|mmcor_desc', names(.))] %>% #drop useless variables
  data.table::as.data.table() %>% # coerce to data.table
  .[actper %in% c(prior, current), ] %>% #keep only dates in prior or current
  .[, `:=`(col = ifelse(actper %in% current, 'current', 'prior'))] %>% # create binary col variable
  .[, actper := NULL] %>% # drop old date variable
  .[, lapply(.SD, sum), by = .(col, product, mmcor_desc = category)] %>% # aggregate
  merge(fridays, by = 'col') #merge with fridays
  
df[, `:=`(avg_cost_per_week = formattable::currency(paid/fridays),
          avg_claims_per_week = formattable::accounting(util/fridays),
          cost_per_claim = formattable::currency(paid/util))]
# create new variables of interest

df <- fidelis::add_var_cols2(df, vars = c('avg_cost_per_week', 'avg_claims_per_week', 'cost_per_claim'))
# convert to 3612 style with this fidelis function
```

With our df data.table ready, I find the easiest way to loop through each product is to make a function that outputs the final table, with the product as an input.

```{r}
display_table <- function(i){
  
  D <- df[product == i, -'product']
  #keep only specified product and drop that variable
  
  D %>% 
    .[order(-avg_cost_per_week_var)] %>% #order in descending avg_cost_per_week_var
    dplyr::mutate(
      avg_cost_per_week_prior = color_bar('lightblue')(avg_cost_per_week_prior),
      avg_cost_per_week_current = color_bar('lightblue')(avg_cost_per_week_current), #note: color_bar returns a function
      avg_cost_per_week_var = red_green_tiles.r(avg_cost_per_week_var),
      avg_cost_per_week_var_pct = red_green_text.r(avg_cost_per_week_var_pct)
      ) %>% # add color formatting
    knitr::kable(
      escape = F, #necessary for use with formattable
      format = 'html', #html output
      table.attr = "style = \"color: black;\"", #black text instead of grey
      col.names = c('COS', rep(c('Prior', 'Current', 'Var', 'Var (%)'), 3)), #lower columnn names
      align = c(rep('l', 1), rep('r', ncol(D)-1)) #specify alignment
      ) %>% # kable is part of the knitr package
    kableExtra::kable_styling(
      full_width = T, #make the table fill the page hoirzontally, probably best as there are 12 data columns
      protect_latex = F, #seems useful for formattable coloring
      position = "center", #center table
      bootstrap_options = c('condensed', 'hover') #i like condense as it saves space and hover makes it easy to read
      ) %>% #kableExtra package is an extension to knitr's kable
    kableExtra::column_spec(1, bold = T) %>% # bold the first column
    kableExtra::add_header_above(c(' ' = 1,
                       'Cost/ Week' = 4,
                       'Claims/ Week' = 4,
                       'Cost/ Claim' = 4)) %>% # add header above with the true column names
    htmltools::knit_print.html() # very important, to be used with Rmarkdown chunk option: results='asis'
  
}
```

Now, with `results='asis'` we can write plain html to be rendered. We must use cat when writing text or tables, and use the `br()` function from the fidelis package for line breaks. Note: graphs can be easily looped with ggplot but plotly objects are problematic and looping is not as easy. For example, the drilldown files I made use plotly but there actually isn't a loop involved, but rather good-ole copy and paste!<br>

This code though will display a table for each product and dump each into a new tabbed section.

## {.tabset .tabset-fade}

```{r results='asis'}
for(i in lob[1:3]){# limit to first three LOB to keep document short
  
  fidelis::br(2)
    cat('###', i)
  fidelis::br(2)
    cat(display_table(i))
  fidelis::br()
}
```

##

